---
title: "Intermediate_R_week3"
author: "Allie Warren"
date: "2024-10-08"
output: html_document
---

Missing Data

Notes:

This notebook covers some of the material from week 3 of the Intermediate R Datacamp track. It also includes some info that isn't covered in the course, including some functions I recommend.

- This notebook uses data on Percentage of Chronic Conditions among Fee-for-Service Medicare Beneficiaries, Washington State and Counties, 2007-2018, from CMS Medicare Chronic Conditions data
  - Data source: https://data.wa.gov/Health/Percentage-of-Chronic-Conditions-among-Fee-for-Ser/qb7g-hu6x/about_data

- Notes: course textbook resource: https://psyteachr.github.io/ads-v3/
 
# Loading Packages 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load packages
library(tidyverse)
# might need to install the package first
# install.packages('naniar')
# naniar is a package for summarizing and handling missing values in R
library(naniar)
# might need to install the package first
# install.packages('simputation')
# simputation is a package for imputing data
library(simputation)

```

# Load Data

Data is on the prevalence of different chronic conditions (Alzheimers, Arthritis, Asthma, Atrial Fibrillation, Autism, COPD, Cancer, Chronic Kidney Disease, Depression, Diabetes, HIV/AIDS, Heart Failure, Hepatitis, Hyperlipedemia, Hypertension, Ischemic Heart Disease, Osteoporosis, Schizophrenia, and Strokes) among Medicare Beneficiaries in Washington counties per year. 

```{r}

# read in data
# for the base file path I am getting the filepath to the folder of the folder that this script is saved in
# then adding the folder data to the file path, as the data files for this repo are stored there
base_path <- file.path(dirname(dirname(rstudioapi::getSourceEditorContext()$path)), "data")
medicare_file <- "Percentage_of_Chronic_Conditions_Medicare_Beneficiaries__Washington.csv"

med <- read_csv(file.path(base_path, medicare_file))

head(med)

```

# Summarizing Missing Data

Functions to summarize the missing data per variable and overall

```{r}
# Summary of missing data using function from the naniar package
missing_med_data <- miss_var_summary(med)
missing_med_data

# Searching for various types of missing data
# looking for missing data that is represented by: . or NA (as a string) or -99
alt_missing_med_data  <- naniar::miss_scan_count(med, c("^\\.$", "NA", "-99"))
alt_missing_med_data
```

# Visualizing Missing Data

```{r}
# plot of the number of missing values per column
gg_miss_var(med)
```


# Replacing Missing Values

```{r}
# replace non standard missing values with NA
# for the column State and County FIPS Code replace any . with NA
med_cleaned <- replace_with_na(med,
               replace = list(`State and County FIPS Code` = "."))

```

# Visualizing Missing Data per Group

```{r}

# subsetting data and simplifying column names for visualization
med_subset <- med_cleaned %>%
  select(County, FIPS = `State and County FIPS Code`, HIV_AIDS = `HIV/AIDS Prevalence (%)`, ASD =  `Autism Spectrum Disorders Prevalence (%)`, HepBC = `Hepatitis (Chronic Viral B & C) Prevalence (%)`, Psych = `Schizophrenia/Other Psychotic Disorders Prevalence (%)`, Year)

# plot missing data per column by County
gg_miss_var(med_subset %>% select(-Year), facet = County)

# plot missing data per column by Year
gg_miss_var(med_subset %>% select(-County), facet = Year)

```


# Missing Data and Relationships between Variables

```{r}

# look at the relationship between COPD prevalence and Hypertension prevalence
# color points by whether that location/year is missing data on stroke prevalence
ggplot(med_cleaned, aes(`COPD Prevalence (%)`,  `Hypertension Prevalence (%)`, color = is.na(`Stroke Prevalence (%)`))) +
  geom_point() + theme_minimal() + labs(color = 'Stroke Prevalence Missing')

# look at the relationship between Hypertension prevalence and Stroke prevalence for each location/year
# (missing values in Stroke Prevalence will not be plotted)
ggplot(med_cleaned, aes(`Hypertension Prevalence (%)`, `Stroke Prevalence (%)`)) +
  geom_point() + theme_minimal()
```

# Imputing Values

Using linear regression imputation methods to impute values for Stroke prevalence based on the
relationship between stroke prevalence and other chronic conditions

```{r}
# mark which rows are NA to keep record of after imputation
med_cleaned <- med_cleaned %>% mutate(Stroke_Flag = is.na(`Stroke Prevalence (%)`))

# select subset of data and simplify column names
med_cleaned_heart <- med_cleaned %>%
  select(Stroke_Flag, Stroke_Prev = `Stroke Prevalence (%)`, Diabetes = `Diabetes Prevalence (%)`, 
         Hypertension = `Hypertension Prevalence (%)`,
         COPD = `COPD Prevalence (%)`, Ischemic_Heart = `Ischemic Heart Disease Prevalence (%)`, County, Year)

# impute values for stroke prevalence 
# this assumes a linear relationship between the stroke prevalence and diabetes, hypertension, COPD, and ischemic heart disease prevalence and county
# then plot the value, highlighting values that were imputed
med_cleaned_heart %>%
  impute_lm(Stroke_Prev ~  Diabetes + Hypertension + COPD + Ischemic_Heart + County) %>%
  arrange(Stroke_Flag) %>%
  ggplot(aes(x = Ischemic_Heart,
             y = Stroke_Prev,
             color = Stroke_Flag, size = Stroke_Flag)) + 
  scale_size_manual(values = c(2, 3)) +
  geom_point() +
  theme_bw() +
  labs(color = 'Stroke Missing Data', size = 'Stroke Missing Data')
```
# Closer Look at the Linear Model

Looking at various linear model for estimating stroke prevalence to evaluate how well we expect the imputation method 
is doing at estimating missing values

```{r}

# R-Squared: how much of the variance in the dependent variable is explained by the independent variables. The greater the R-Squared, the more predictive power a model has (0 to 1)
# Coefficient significance: p-value per coefficient
# Residual standard error (RSE): the average amount that the data points deviate from the regression line. The less deviation there is, the more closely the model matches the data.


lin_model <- lm(Stroke_Prev ~ Diabetes + Hypertension + COPD + Ischemic_Heart + County, data = med_cleaned_heart)
summary(lin_model)


# alternative model
alt_lin_model <- lm(Stroke_Prev ~ COPD + Ischemic_Heart + Year + County, data = med_cleaned_heart)
summary(alt_lin_model)
```