---
title: "Intermediate_R_plotting"
author: "Allie Warren"
date: "2024-10-10"
output: html_document
---

Plotting In R

Notes:

This notebook covers some of the material from month 4 of the Intermediate R DataCamp track, and some info that isn't covered in the course, on creating different types of plots in R.

Topics covered:
- creating a scatter plot
- labeling points w/ ggrepel
- creating bar plots - rotating axis text, using position_dodge
- creating a density plot
- creating a violin plot with jittered points
- creating a WA state map plot
- creating alternate color schemes - viridis or custom
  - R color names: https://r-charts.com/colors/
  - viridis colors: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
  - generating accessible colors: https://venngage.com/tools/accessible-color-palette-generator
- plotting themes
  - publication ready themes: https://www.datanovia.com/en/blog/tag/ggpubr/

This notebook uses data on respiratory hospitalizations for WA state (https://doh.wa.gov/data-and-statistical-reports/diseases-and-chronic-conditions/communicable-disease-surveillance-data/respiratory-illness-data-dashboard) and an example of line level data on ebola cases. 
 
# Loading Packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#This line of code installs the pacman page if you do not have it installed - if you do, it simply loads the package
if(!require(pacman))install.packages("pacman")
# then we can use the p_load function to load or, as needed, install packages
# similar to the conditional statements we used in previous notebooks
pacman::p_load('tidyverse', 'ggpubr', 'ggrepel', 'readxl', 'viridis', 'ggridges', 'sf')


```


# Loading Data

Two example datasets
  - Respiratory hospitalizations and emergency room visits for Washington 
  - ebola case line level data

```{r}

# for the base file path I am getting the file path to the folder that this script is saved in
# if you has stored the data in a different folder, then you will need to change the base path here
base_path <- file.path(dirname(dirname(rstudioapi::getSourceEditorContext()$path)), "data")

# WA state respiratory disease hospitalization rates
resp_hosp <- read_csv(file.path(base_path, "Respiratory_Disease_RHINO_Downloadable_Data.csv"))
# data from a paper on ebola cases - has per individual case data
ebola <- readxl::read_xlsx(file.path(base_path, "Finaldataset F1000.xlsx"))

# View the data
head(resp_hosp)

head(ebola)

```


# Scatter Plot of Hospitalizations over Time

Creating a scatter plots with different point colors, shapes, and sizes per category

```{r}
# using colors to mark the different respiratory illnesses and shape and size to mark emergency visits vs hospitalizations
# filter to all demographics and statewide data
filter(resp_hosp, `Demographic Category` == 'Overall', Location == 'Statewide') %>%
# plot percent per week, with COVID-19 marked in red, Flu marked in green, and RSV marked in blue
# and emergency visits denoted with a circle and hospitalizations with a triangle
ggplot(aes(`Week Start`, `1-Week Percent`, color = `Respiratory Illness Category`, shape = `Care Type`, size= `Care Type`)) +
  # set slightly different size points for hospitalizations vs emergency visits to further differentiate
  scale_size_manual(values = c(1.6, 2.3)) +
  # set opacity to .8 to make it easier to see overlapping points
  geom_point(alpha = 0.8) +
  # set theme to minimal to make the plot cleaner
  theme_minimal() +
  # rename label for color in the legend to shorten
  labs(color = 'Illness') +
  # add title
  ggtitle('Respiratory Illness Hospitalizations and Emergency Visits')



```
# Line Plot

Creating a similar plot as asbove, but as a line plot instead of a scatter plot
Using different colors to denote illness, and different line types to denote Hospitalizations vs Emergency Visits
```{r}
filter(resp_hosp, `Demographic Category` == 'Overall', Location == 'Statewide') %>%
ggplot(aes(`Week Start`, `1-Week Percent`, color = `Respiratory Illness Category`, linetype = `Care Type`)) +
  # set plot type to line and set line thickness
  geom_line(linewidth=1.2) +
    # change theme to minimal
  theme_minimal() +
  # rename label for color in the legend to shorten
  labs(color = 'Illness') +
  # add title
  ggtitle('Respiratory Illness Hospitalizations and Emergency Visits')


```

# Scatter Plot with Labels

Show two plots, one using default label positioning, and one using geom_text_repel to position labels. geom_text_repel is a 
function from the ggrepel package that creates 'repulsive' text labels so that the text labels do not
overlap each other

```{r}

# filter respiratory data to overall demographics and COVID-19 and week 36 data
# then we convert the data from long format to wide format so that we have a separate column for hospitalizations and emergency room percentage
week6_covid_data <- resp_hosp %>% filter(`Demographic Category` == 'Overall', `Respiratory Illness Category` == 'COVID-19', `Week Start` == "2024-09-01") %>%
  pivot_wider(names_from = `Care Type`,
    values_from = c(`1-Week Percent`)) 


# then we plot hospitalizations vs emergency visits for each health region in the state
# and add labels to the points denoting the health region (this makes it easier to identify which point represents
# each region instead of using a legend)
overlapping_labels <- week6_covid_data%>%
  ggplot(aes(Hospitalizations, `Emergency Visits`, color = Location)) +
  geom_point() + 
  theme_bw() +
  # remove the legend, as we are labeling the points instead
  theme(legend.position = 'none') +
  # add text labels
  ggplot2::geom_text(aes(x = Hospitalizations, 
                      y = `Emergency Visits`, 
                      label = Location)) +
  # add title
  ggtitle('COVID-19 1 Week Percent for 2024-09-01 (default label positions)')

overlapping_labels

# using ggepel::geom_text_repel instead of just geom_text to label points so labels don't overlap
well_labeled <- week6_covid_data %>%
  ggplot(aes(Hospitalizations, `Emergency Visits`, color = Location)) +
  geom_point() + 
  theme_bw() +
  # remove legend
  theme(legend.position = 'none') +
  # use geom_text_repel to label points
  ggrepel::geom_text_repel(aes(x = Hospitalizations, 
                      y = `Emergency Visits`, 
                      label = Location)) +
  # add title
  ggtitle('COVID-19 1 Week Percent for 2024-09-01 (with adjusted labels)')

well_labeled


```

# Bar Plots - example with Ebola data
- geom_bar takes in one input and creates bars of height proportional to the number of cases in each group
- geom_col (or geom_bar(stat = 'identity')) takes x and y inputs and creates bars of height equal to the y value for each group 

```{r}
# creating a simple bar plot of counts per group
ebola_simple_bar_plot <- ggplot(ebola) + 
  geom_bar(aes(`Age categories`)) +
  theme_bw() +
  xlab('Age') +
  ggtitle('Ebola Cases by Age (default)')
ebola_simple_bar_plot 

# creating the same plot, but using factors to set the order of variables for plotting
# by default the x-axis will be ordered alphabetically, with may not accurately represent the ordering of our data
# by making the variable a factor we can add levels and set the ordering of the variables
ebola <- ebola %>%
  mutate(Age_Groups = factor(`Age categories`, levels = c('4 and below', '5 to 14 years', '15 to 29 years', '30 to 44 years', '45 to 59 years', '60 and above', NA)))

ebola_ordered_bar_plot <- ggplot(ebola) +
  geom_bar(aes(Age_Groups)) +
  theme_bw() +
  xlab('Age') +
  ggtitle('Ebola Cases by Age (w/ ordering of x-axis)')
ebola_ordered_bar_plot

# creating a multi variable bar plot
ebola_multivar_bar_plot <- ggplot(ebola) +
  # adding a grouping variable (asthenia weakness) to create a mutli variable bar plot
  # we add position_dodge() so that the bars are next to each other, instead of stacked
  geom_bar(aes(Age_Groups, fill = asthenia_weakness), position = position_dodge()) +
  # using theme from ggpubr for better visualizatin
  theme_pubclean() +
  # adjust axis text to be angled so that it fits in display
  theme(axis.text.x = element_text(angle = 54, hjust=1)) +
  # label x axis
  xlab('Age') +
  # adjust label of th legend
  labs(fill = 'Symptom of Asthenia Weakness') +
  # add title
  ggtitle('Ebola Cases by Age and Symptom')
ebola_multivar_bar_plot


```

# Bar plots - with labels
- here we use geom_col to create bars of a specified height
- and we create a custom ordering of the x-axis and custom colors
```{r}
# filter hospitalization data to get RSV hospitalizations overall for a specific week per location
rsv_hosp <- filter(resp_hosp, `Week Start`== "2022-11-20", `Respiratory Illness Category` == 'RSV', `Demographic Category` == 'Overall', `Care Type` == 'Hospitalizations')

# adjust ordering of x-axis so that it is the statewide value, then each region in order of decreasing value
# we use 'order' to get the order of locations in decreasing value, 
# then use factor to set the levels in the order we determined
ordered_locations <- rsv_hosp$Location[order(rsv_hosp$`1-Week Percent`, decreasing = TRUE)]
adjusted_ordered_locations <- c('Statewide', ordered_locations[ordered_locations != "Statewide"])
rsv_hosp$Location <- factor(rsv_hosp$Location, levels = adjusted_ordered_locations)

rsv_hosp %>%
  ggplot(aes(Location, `1-Week Percent`)) +
  # creating different color bar plot for statewide vs health regions
  geom_col(aes(fill = (Location == 'Statewide'))) +
  # set custom fill color by hex code
  scale_fill_manual(values = c('#92aab9', '#46738b')) +
  # adding text of y axis value to the top of each bar plot
  geom_text(aes(label=`1-Week Percent`), vjust=-0.3) +
  # using theme from ggpubr package
  theme_classic2() +
  # adjusting axis text angle and removing legend
  theme(axis.text.x = element_text(angle = 45, hjust=1), legend.position = 'none') +
  # don't clip plot to extent of the plot panel to try to prevent text labels from being cut off
  coord_cartesian(clip = "off") +
  # add title
  ggtitle('Hospitalization Rate for RSV for 2022-11-20')

```

# Lollipop Plot

We can plot the same data as above in a slightly different format using a lollipop plot
Instead of including the statewide value as a bar in the plot we will add a labeled dashed line to represent it
```{r}
# creating a variable of the statewide percent
statewidelevel <- filter(rsv_hosp, Location == 'Statewide')$`1-Week Percent`
  
# remove statewide from the data, so we can separately add it to the plot
filter(rsv_hosp, Location != 'Statewide') %>%
  ggplot(aes(Location, `1-Week Percent`)) +
  # creating a line segment for each location
  geom_segment(aes(x = Location, xend = Location, y = 0, yend =`1-Week Percent`), size = 1.6, color = '#92aab9') +
  # adding a circle to the end of each line segment
  geom_point(size=6.5, fill="#713925", color = 'black', pch = 21) +
  # adding text of the percent to end of lollipop
  geom_text(aes(label = `1-Week Percent`), color = "white", size = 3) +
  # add a dashed line at the location of the Statewide percent 
  geom_hline(yintercept=statewidelevel, linetype='dotted', col = '#46738b', size = 1.1) +
  # labeling the line (with slight offset and at selected location) on the plot
  annotate("text", x = "Olympic Community of Health", y = (statewidelevel + .7), label = "Statewide Percent", size = 3.5) +
  # using theme from ggpubr package
  theme_classic2() +
  # flip coordinates to change orientation of the plot
  coord_flip() +
  # remove legend
  theme(legend.position = 'none') +
  # add title
  ggtitle('Hospitalization Rate for RSV for 2022-11-20')
```

# Overplotting
We can use point transparency or plot the density of points in a plot to make it easier
to visualize dense plots
```{r}
# filter to covid hospitalizations per ACH region across all weeks
covid_hosp <- filter(resp_hosp, Location != 'Statewide' & `Demographic Category` == 'Overall' & `Care Type`=='Hospitalizations' & `Respiratory Illness Category` == 'COVID-19')

# can make points more transparent (lower alpha value) to make it easier to see overlapping points
ggplot(covid_hosp, aes(Week, `1-Week Percent`)) +
  geom_point(alpha = 0.2) +
  theme_minimal() +
  ggtitle('COVID Hospitalizations')

# can also bin the points into rectangles and map the density of the points
ggplot(covid_hosp, aes(Week, `1-Week Percent`)) +
  stat_bin2d(bins = 70) +
  # we use a color schema from the package viridis to visualize the density
  scale_fill_viridis() +
  theme_minimal() +
  ggtitle('COVID Hospitalizations')


```

Density Plot - example with Ebola cases

```{r}
# creating a density plot of ages per vaccination status
ebola_density_plot <- ggplot(ebola) +
  # set lower opacity (0.5) to make it possible to view the different groups
  geom_density(aes(age_in_years, fill = `Vaccination status`), alpha = 0.5) +
  # using theme from ggpubr package
  theme_classic2() +
  xlab('Age') +
  ggtitle('Distribution of Ebola Cases by Age and Vaccination Status')

ebola_density_plot


# Another approach for visualizing density plots for multiple categories
# ggridges package has a lot of cool possibilities for creating density plots: https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html
ebola_density_plot_v2 <- ggplot(ebola) +
  # use the geom_density_ridges function from the ggridges package
  # the x value stays the same, and we use our grouping variable for the y-axis
  geom_density_ridges(aes(x = age_in_years, y = `Vaccination status`, fill = `Vaccination status`), alpha = 0.9) +
  # using theme from ggpubr package
  theme_classic2() +
  # alternate colors of the density ridges to make visualization easier
  scale_fill_cyclical(values = c("#4040B0", "#9090F0")) +
  xlab('Age') +
  ggtitle('Distribution of Ebola Cases by Age and Vaccination Status') +
  # remove legend (as categories are explained on the y-axis)
  # adjust the positioning of text on the y axis to make it clearer which plot aligns with which label
  theme(legend.position = 'none', axis.text.y = element_text(vjust=-2)) +
  # truncating plot at 0, as values less than 0 don't make sense in this context
  xlim(0,max(ebola$age_in_years, na.rm = TRUE) + 10)


ebola_density_plot_v2
```


Box plot vs Violin plot 

Visualizing the distribution of a continuous variable - a boxplot is a common plot type for visualizing this data, and also incorporates five summary statistics (median, two hinges, and whiskers) as well as outlying points. A violin plot is another option, which blends a box plot and a density plot. We can compare these two types of plots, and also show how to create them and include the points and use a custom color palette
```{r}
# for EVD status confirmed cases, plot distribution of ages per vaccination status
ggplot(filter(ebola, EVD_status == 'Confirmed'), aes(x = `Vaccination status`, y = age_in_years, fill = `Vaccination status`)) +
  geom_boxplot(alpha = 0.5) +
  # adding points to plot, and using jitter to randomly move the points sightly so that they don't overlap as much
  geom_point(position = position_jitter(seed = 1, width = 0.2)) +
  # setting a custom color palette
  scale_fill_manual(values = c("steelblue4", "gold2", "seagreen", "hotpink3")) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none") + 
  ylab('Age') +
  ggtitle('Age Distribution per Vaccination Status (box plot)')

# for EVD status confirmed cases, plot distribution of ages per vaccination status
ggplot(filter(ebola, EVD_status == 'Confirmed'), aes(x = `Vaccination status`, y = age_in_years, fill = `Vaccination status`)) +
  geom_violin(alpha = 0.5) +
  # adding points to plot, and using jitter to randomly move the points sightly so that they don't overlap as much
  geom_point(position = position_jitter(seed = 1, width = 0.2)) +
  # setting a custom color palette
  scale_fill_manual(values = c("steelblue4", "gold2", "seagreen", "hotpink3")) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none") + 
  ylab('Age') +
  ggtitle('Age Distribution per Vaccination Status (violin plot)')


```


# Washington State Map Plot
- using ggplot to create a map plot of WA state counties
- get the center of each county to place labels of the county name
- fill in each county with COVID-19 hospitalization rate using an alternate color palette 

```{r}

# hospitalization data contains ACH regions rather than counties
# we can create a mapping of ACH to counties for mapping purposes
ach_mapping <-
  tibble(ACH = c('Olympic Community of Health', 'North Sound', 'Healthier Here', 'Elevate Health', 'Cascade Pacific Action Alliance', 'Southwest Washington', 'Greater Health Now', 'Thriving Together NCW', 'Better Health Together'),
             counties = list(c('Clallam', 'Jefferson', 'Kitsap'), 
                          c('Whatcom', 'Skagit', 'Snohomish', 'San Juan', 'Island'),
                          c('King'),
                          c('Pierce'),
                          c('Grays Harbor', 'Mason', 'Thurston', 'Pacific', 'Lewis', 'Wahkiakum', 'Cowlitz'),
                          c('Clark', 'Skamania', 'Klickitat'),
                          c('Kittitas', 'Yakima', 'Benton', 'Franklin', 'Walla Walla', 'Columbia', 'Garfield', 'Asotin', 'Whitman'),
                          c('Okanogan', 'Chelan', 'Douglas', 'Grant'),
                          c('Adams', 'Lincoln', 'Ferry', 'Stevens', 'Pend Oreille', 'Spokane')))



# import the county map data (this is available from the ggplot2 package)
counties <- map_data("county")
# covert county names to title format so they match with our data
counties$subregion <- str_to_title(counties$subregion)

# filter to WA state
wa_counties <- filter(counties, region == 'washington')

# get center of each county for plot labeling purposes
cntrd <- function(x) {
  data.frame(geosphere::centroid(as.matrix(x[,c("long", "lat")])))
}
centroid_list <- group_by(wa_counties, subregion) %>%
  do(cntrd(.))
# create table of centroid coordinates and the name of the county for labeling the plot
centroid_list <- data.frame(long = centroid_list$lon, lat= centroid_list$lat, name = centroid_list$subregion)



# filter hospitalizations data to a specified respiratory condition (COVID-19, Flu, or RSV),
# care type (Hospitalizations or Emergency Visits) and week start
illness <- 'COVID-19'
filter_date <- '2024-09-22'
cur_care_type <- 'Hospitalizations'
resp_hosp_regions <- filter(resp_hosp, Location != 'Statewide' & `Demographic Category` == 'Overall' & `Respiratory Illness Category` == illness & `Week Start` == filter_date & `Care Type` == cur_care_type & !is.na(`1-Week Percent`))

# join hospitalization data with WA mapping data
resp_hosp_regions <- left_join(tibble(resp_hosp_regions), ach_mapping, by = c('Location' = 'ACH'))
resp_hosp_regions <- resp_hosp_regions  %>%
  # expand the column containing the list of counties into separate rows
  # so there is one row per county, instead of one row per ACH region
  unnest(counties)
resp_hosp_regions_wa <- left_join(wa_counties, resp_hosp_regions, by = c('subregion' = 'counties'))


# Plot map
wa_map_plot <- ggplot(data = resp_hosp_regions_wa) +
  # use viridis color scale - good for creating a visually accommodating distinct color scale
  scale_fill_viridis_c() +
  # plot lat and longitude outline of each county to create map plot and fill in county with hospitalization data
  # outline states with white
  geom_polygon(aes(x = long, y = lat, group = group, fill = `1-Week Percent`), color = 'white') +
  # add label for each county name at center of county
  geom_text(data = centroid_list,
                       aes(label= name, x = long, y = lat), size = 3, color = 'black') +
  # set simple theme
  theme_bw() +
  # add title based on data filtered
  ggtitle(paste(illness, cur_care_type, "for", filter_date)) 

wa_map_plot

```
# Alternative Process for Creating Washington State Map Plot
- using ggplot and the sf package to create a map plot of WA state counties
- uses more built in functions
- map looks slightly different for island counties

```{r}
# pull US mapping data
us_county_map <- usmap::us_map('counties')
# subset to WA state data
WA_map_counties <- filter(us_county_map, abbr == 'WA') %>%
  # simplify county names
	mutate(county = str_replace(county, " County", ""))
# convert a geometry object to an sfc object for plotting purposes
WA_sfc <- st_as_sfc(WA_map_counties, crs = usmap_csv()@projargs)
# get subset of map object
# creates sf object, which extends data.frame like objects with a  simple feature list column
WA_sf_data <- st_sf(data.frame(fips = unique(WA_map_counties$fips), county = WA_map_counties$county, geometry = WA_sfc))
# join map data with the data you want to plot/fill in each county with
WA_sf_map <- left_join(WA_sf_data, resp_hosp_regions, by = c('county' = 'counties'))
# find the centroid of each county
WA_sf_map$centroids <- st_centroid(WA_sf_map$geometry)

wa_map_plot2 <- ggplot(WA_sf_map) +
  # use geom_sf to create map - pulls from geometry column in the data
  # set fill for each county
	geom_sf(aes(fill =`1-Week Percent`), color = 'white') +
  # change fill - alternate color palette
  scale_fill_viridis_c(option = 'plasma') +
  # add county labels
	geom_sf_text(aes(label = county), size = 3, color = 'black') +
  coord_sf(crs = st_crs(4283)) + #needed to make flat/horizontally aligned map for visualizing purposes
  # set to simple theme
  theme_bw() +
  # remove axis labels
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  # remove axis labels
  xlab("") + ylab("") +
  # add title based on data filtered
  ggtitle(paste(illness, cur_care_type, "for", filter_date)) 

wa_map_plot2

# save a plot
# optional can set the size of the plot
ggsave(file.path(base_path, "wa_resp_hospitalizations_map_plot.png"), width = 10, height = 8)
```