---
title: "Intermediate_R_week1"
author: "Allie Warren"
date: "2024-09-27"
output: html_document
---

Notes:

This notebook covers some of the material from course 1 of the Intermediate R datacamp track
It also includes some info that isn't covered in the course, including some functions I recommend and
a function I wrote to pull and format ACS data using their API.

Useful features within R Studio:
- use install.packages('name_of_package') to install a package, then library(name_of_package) to load it
- use View(df) to pull up a table that you want to look at. If your table is very large this might be slow to load,
use View(head(df, n = 100)) to view just the first 100 (or however many you specify) rows of the table
- use ?, such as ?union, in the console to pull up help info about how a function works

The 'Knit' button in the top panel will run all the code and create a HTML file, with text/plots/data embedded 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# load packages - if you don't already have these installed you will need to run install.packages(c('tidyverse', 'httr', 'lubridate'))

library(tidyverse)
# for fetching ACS data using API 
library(httr)
# for working with dates
library(lubridate)
```


# Helper function for pulling ACS data

This function uses concepts not covered in this course, but can be used to pull and format data from
ACS using their API. It is used to get example datasets for demonstrating the concepts of the course
on real world data.

The tidycensus package can also be used to pull ACS data: 
- https://walker-data.com/census-r/an-introduction-to-tidycensus.html
- requires getting an API key from: https://api.census.gov/data/key_signup.html

```{r}

# function to pull ACS tables using the API
# inputs:
# year: (string) year of data table
# survey_type: (string) acs5 or acs1
# group_number: (string) table number
# location: (string) survey location in ACS format
# returns: tibble of ACS data
fetch_acs <- function(year, survey_type, group_number, location) {
  # API path to specified table
  acs_url <- file.path("https://api.census.gov/data", year,"acs", survey_type,
                     paste0("subject?get=group(", group_number,
                            ")&ucgid=pseudo(", location, ")"))
  # fetch the table
  response <- httr::GET(acs_url, httr::timeout(10))
  acs_response <- httr::content(response)
  # convert it to a formatted dataframe
  out_df <- as.data.frame(matrix(data = NA, nrow = (length(acs_response) - 1),
                                 ncol = length(acs_response[[1]])))
  # first item in response is column names
  colnames(out_df) <- unlist(acs_response[[1]])
  # loop through the other items in response to populate the data frame
  for(i in 2:length(acs_response)) {
    cur_row <- unlist(lapply(acs_response[[i]], function(x) ifelse(is.null(x[[1]]), NA, x)))
    out_df[(i-1),] <- cur_row
  }
  # return the table as a tibble
  return(as.tibble(out_df))

  
}

```

# Load Data
```{r}

# Idaho and Washington ACS 5 year 2022 education data
# "S1501" specifies the ACS tables on EDUCATIONAL ATTAINMENT
# "0400000US53$0500000" specifies WA, "0400000US16$0500000" specifies Idaho
education_WA <- fetch_acs("2022", "acs5", "S1501", "0400000US53$0500000")
education_Idaho <- fetch_acs("2022", "acs5", "S1501", "0400000US16$0500000")

# Washington ACS 5 year 2022 data on poverty rates
# "S1701" specifies the ACS table on POVERTY STATUS IN THE PAST 12 MONTHS
poverty_rate <- fetch_acs("2022", "acs5", "S1701", "0400000US53$0500000") 

# example viewing the table
View(education_WA)
```

# Combine Data
```{r}

# combine WA and Idaho data - union combines all rows in the two dataframes excluding duplicates (in this case it will concatenate the two tables)
# it concatenates columns with the same names - it will fail if columns are not of the same type
# (for example string in one table and numeric in the other) or if a column exists in only one data frame
education <- union(education_WA, education_Idaho)

```

# Select & Tidy Evaluation
- selecting columns
- !! and !!!
- enquo()
- {{}}
- :=

For more examples/explanations: https://www.r-bloggers.com/2019/12/practical-tidy-evaluation/

```{r}
# selecting columns

# column names
common_col <- "NAME"
# columns in poverty_rate table
totalpop_determined <- c('S1701_C01_001E', 'S1701_C01_001M') # total population and margin of error
numbelowpoverty = c('S1701_C02_001E', 'S1701_C02_001M') # number below poverty rate and margin of error
# columns in education table
popover25 = c('S1501_C01_006E', 'S1501_C01_006M') # total population over 25 and margin of error
numover25bachelors <- c('S1501_C01_015E', 'S1501_C01_015M') # number over 25 with a bachelors degree and margin of error

# select relevant variables using {{}} or !!
# {{}} and !!enquo() do (mostly) the same thing, {{}} is just cleaner to write


# for unquoting and unquote-splicing we can use !! and !!! respectively
# !! for single variables (like common_col in this example) and !!! for vectors (lists)
poverty_rate_subset <- poverty_rate %>%
  select(!!common_col, !!!totalpop_determined, !!!numbelowpoverty)

# {{}} will also extract a stored variable name 
education_subset <- education %>%
  select({{common_col}}, {{popover25}}, {{numover25bachelors}})


# using functions to select columns that match a pattern
# selecting columns that start with S1501_C01_006 - will select both the columns S1501_C01_006E and S1501_C01_006M
education_subset_namea <- education %>% 
  select(starts_with('S1501_C01_006'))

# does this same thing as the function above, but just using the syntax ^, which means start of the string
# the matches functions allows you can uses regular expressions
education_subset_nameb <- education %>% 
  select(matches('^S1501_C01_006'))

# similarly ends_with and matches('string$') do the same thing
```

```{r}

# using enquo, !!, and {{}} with functions 

# here we are using enquo to capture a bare column name (not quoted)
# it will convert the bare column names into expressions which we then use !! to understand
select_by_column <- function(df, column_var) {
  df %>% select(!!enquo(column_var))
}

# here were are just using !! because for this function we will pass a quoted column name
# and !! unquotes it 
select_by_column_name <- function(df, column_name) {
  df %>% select(!!column_name)
}

# here we are using {{}} which can do the same thing as enquo
# it can handle quoted and unquoted expressions and is more concise
select_by_column_all <- function(df, col) {
  df %>% select({{col}})
}

# for select_by_column using enquo we can pass the column name as a symbol (not as a quoted string) and it will evaluate it
education_subset_test1 <- select_by_column(education, NAME)
# for select_by_column_name we pass the column name as a quoted string, which !! unquotes
education_subset_test2a <- select_by_column_name(education, "NAME")
# or we can pass it as a variable of a quoted string
col_name <- "NAME"
education_subset_test2b <- select_by_column_name(education, col_name)
# for select_by_column_all we use {{}} which can handle many different cases
education_subset_test3a <- select_by_column_all(education, NAME)
education_subset_test3b <- select_by_column_all(education, "NAME")
education_subset_test3c <- select_by_column_all(education, c(NAME, S1501_C01_006E))
education_subset_test3d <- select_by_column_all(education, numover25bachelors)

```

```{r}
# using the walrus operator :=
# this is used for creating or renaming columns dynamically

bachelors_rate_col <- paste0(numover25bachelors[1], "_rate")
  
# Using := for dynamic column creation
# I use !! to unquote the column name
# and := to create a column of the name (instead of the name 'bachelors_rate_col')
# I also need to use !!sym() for the calculation because:
# When working with dplyr operations on columns, we need symbols not strings (vs select works with strings or variables)
# Using just !! would unquote to: "S1501_C01_015E" (still a string)
# so dplyr would look for a column literally named "S1501_C01_015E" (with quotes)
# using both !!sym(numover25bachelors[1])  
# First converts "S1501_C01_015E" to a symbol, then unquotes to: S1501_C01_015E (now a symbol)
# so that dplyr looks for a column named S1501_C01_015E (no quotes)
# We are using sym() rather than enquo() because we are passing a string not a bare column name

education_ratio_a <- education %>%
  # select relevant columns using same process as above
  select({{common_col}}, {{popover25}}, {{numover25bachelors}}) %>%
  mutate(
    # Create bachelor's degree percentage
    !!bachelors_rate_col := (as.numeric(!!sym(numover25bachelors[1])) / as.numeric(!!sym(popover25[1]))) * 100)


# example of not using !! and :=
# here the column will be named bachelors_rate_col instead of "S1501_C01_015E_rate"
education_ratio_b <- education %>%
 select({{common_col}}, {{popover25}}, {{numover25bachelors}}) %>%
  mutate(
    bachelors_rate_col := (as.numeric(!!sym(numover25bachelors[1])) / as.numeric(!!sym(popover25[1]))) * 100)

# without :=
# here the column will be named bachelors_rate_col instead of "S1501_C01_015E_rate"
education_ratio_c <- education %>%
 select({{common_col}}, {{popover25}}, {{numover25bachelors}}) %>%
  mutate(
    # Create bachelor's degree percentage
   bachelors_rate_col = (as.numeric(!!sym(numover25bachelors[1])) / as.numeric(!!sym(popover25[1]))) * 100)

```

# Table Joins

```{r}

# join the two tables
# full join includes all rows from each dataframe, even if there isn't a match in the other df
# we are joining on the column NAME (could also join on multiple columns or columns of differing names)
# for example if in education_subset it was named County and in poverty_rate_subset it was NAME
# then we would use by = c('County' = 'NAME') 
acs_combined <- full_join(education_subset, poverty_rate_subset, by = 'NAME')


# Example of joining the tables if the column you are joining on differs in name between the two tables
# I am creating an example table where I changed the county name column from NAME to County
ex_education <- education_subset %>%
  rename(County = NAME)

# now joining that new table with the previous one 
# use the by = c('County' = 'NAME') to specify that in the first table the column
# is named County and in the second table the column is named NAME
ex_acs_combined <- full_join(ex_education, poverty_rate_subset, by = c('County' = 'NAME'))

```

# Filtering Data

We can use filter to subset our data based on a condition

```{r}

# filter to just WA
acs_combined_WA <- acs_combined %>%
  # I am using str_detect which is a tidyverse function for string matching
  # it returns TRUE/FALSE for whether each input matches, aka contains the substring or pattern, passed as input
  # so that we filter to all rows in the table with Washington in the NAME column
  filter(str_detect(NAME, ", Washington"))

```

# Transforming Data

## Mutate Across

Transform the data across a set of columns 

```{r}
# mutate across - used to transform across a set of columns
# converting all columns across this range of columns to numeric
acs_combined_WA <- acs_combined_WA %>%
  mutate(across(S1501_C01_006E:S1701_C02_001M, as.numeric))

# perform the same mutate across those columns now replacing values that are less than 0 with NA
# the function replace() takes in the 3 arguments: a vector, the index vector (which ones to replace, can be a condition/expression), and the value to replace it with
# the different components of ~replace(., .<0, NA) mean:
# ~ means do this to each column
# . means the current column we're working on
# .<0 means for the current column all values less than 0
# NA - the input were a replacing with, which indicates missing values
acs_combined_WA <- acs_combined_WA %>%
  mutate(across(S1501_C01_006E:S1701_C02_001M,  ~replace(., .<0, NA)))

```

## Mutate - string and value transformation 
```{r}
# str_replace is another tidyverse function for working with strings
# it takes in 3 inputs, the input vector (or column), the pattern that we are matching and replacing, and what we should replace it with
# in this case for the column NAME, we are replacing all text including and following a comma ( . means any symbol and * means any number of those symbols, so it will replace all text following a , ) and removing it (replacing with the empty string) - so this will get rid of ', Washington' or anything following , so we just have the county name
# then we calculate rates and the upper and lower confidence intervals 
acs_combined_WA <- acs_combined_WA %>%
  mutate(County = str_replace(NAME, ",.*", ""),
         percent_bachelors_25_plus = S1501_C01_015E/S1501_C01_006E,
         pb25_ucl = (S1501_C01_015E + S1501_C01_015M)/S1501_C01_006E,
        pb25_lcl = (S1501_C01_015E - S1501_C01_015M)/S1501_C01_006E)
```

# Select and Visualize Data

```{r}


# visualizing data using ggplot - creating a bar plot of the percentage of people over 25
# with a bachelors degree and confidence intervals per county
# pass in the table acs_combined_WA
acs_WA_plot <- acs_combined_WA %>%
  # on the x-axis is the County and on the y-axis is the percentage of people with a bachelors degree
    ggplot(aes(County, percent_bachelors_25_plus)) +
  # set plot type to column (or bar plot) - geom_col is equivalent to geom_bar(stat = 'identity')
    geom_col() +
  # add lines for the margin of error on each value (using the columns we created above) and set some of the aesthetics for those lines
    geom_errorbar( aes(x=County, ymin=pb25_lcl, ymax=pb25_ucl), width=0.25, colour="red", alpha=0.9, size=.8) +
    # set theme to a cleaner visual
    theme_bw() +
    # add title to plot
    ggtitle("Percent of People 25+ with a Bachelors Degree or Higher") +
    # change the y label to be more concise
    ylab("Percent") +
    # rotate axis text so it is easier to visualize
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 

# View the plot
acs_WA_plot

# saving the plot to a file
# I am saving it to the folder above the folder that this script is in, you can also change it
# and I am naming the final plot image example_plot
base_folder <- dirname(dirname(rstudioapi::getSourceEditorContext()$path))
ggsave(file.path(base_folder, "example_plot.png"), acs_WA_plot)
```

# Dates

```{r}
# adding the current date to the table to mark when it was accessed
# using the lubridate package which has lots of functions for working with dates
acs_combined_WA <- acs_combined_WA %>%
  mutate(date_accessed = lubridate::today())


```


# View 

```{r}

# Viewing the first 20 rows of the table
head(acs_combined_WA, n = 20)


```